Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2019-08-25 Manuscript/manuscript.Rmd
edits

## SMOTE RE-SAMPLING COUPLED WITH A DISTANCE-TYPE FEATURE SPACE PROVIDED OPTIMAL CELL LINE SPECIFIC MODEL PERFORMANCE

@@

We first evaluated the least granular cell line specific random forest classifiers [??? Unclear; !!!discuss in meeting], those were ones implemented using OC predictor types, without data re-sampling. We found that these models exhibited poor MCCs across all resolutions (Figure 5 A-B (red)). We then assessed models built on OP predictor types, while still not performing any data resampling. Performances were similar to that of OC types, with no visible improvement across resolution, for either cell line (Figure 5 C-D (red)). This indicated that the added complexity of using overlap percents from the functional genomic annotations did not affect predictive performance. However, when cell line specific classifiers were implemented using distance-type predictors and no re-sampling, we found that there was a significant improvement in MCC compared to the other two predictor types (Figure 5 E-F (red)). We then further explored the impact of data re-sampling on model performance, across cell line and resolution. For cell line specific random forest classifiers built on balanced data, we found that SMOTE re-sampling provided the optimal MCC, consistently outperforming models built on ROS or RUS respectively, across both predictor type and resolution (Figure 5). Specifically, the most improvement was seen when combining SMOTE re-sampling with distance-type predictors for both the GM12878 and K562 cell lines.

## MODEL PERFORMANCE PEAKED AT 10 KB AND DECREASED AS Hi-C DATA RESOLUTION DECREASED

Interestingly, we found that performance peaked at 10 kb resolution and then degraded as data resolution decreased (Figure 5). The reason for the improved performance between 5 kb and 10 kb data resolutions can be attributed to the added sparcity in the interaction frequency matrices at finer resolutions. This sparcity corresponds to inaccurate TAD calling which leads to worse model performance. The difference in performance between 5 kb and 10 kb resolutions for K562 was found to be slightly steaper, indicating that this issue of sparcity was more severe at finer resolutions for K562. Additionally, performances tended to be much more variable between chromosomes at lower resolutions. Moreover, this decreasing trend in performance was much more evident under the optimal combination of data level characteristics (SMOTE resampling with distance type predictors). These results imply that there is an ideal resolution in TAD boundary prediction, and that that an optimally performing cell line specific random forest classifier <!--, one that is able to best discriminate between genomic bins that are near versus far away from TAD boundaries,--> is one that is built on 10 kb resolution data, with distance-type predictors, and incorporates both random under- and over-sampling.

![**SMOTE re-sampling coupled with distance-type predictors yielded optimal MCCs across resolution for cell line specific models.** Performances of random forest classification models built on each of the 3 predictor types including (A-B) OC, (C-D) OP, and (E-F) Distance, and across both the GM12878 and K562 cell lines. Within each cell line, predictor type, and resolution, chromosome-specific training data was re-sampled using different techniques including no re-sampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and *Synthetic Minority Over Sampling TEchnique* (SMOTE; purple). Performances were evaluated using MCCs and aggregated by averaging across each chromosome, with error bars representing 1 standard deviation away from the mean.](Figures/model_performances_v3.png)

>>>>

## RANDOM FOREST CLASSIFIERS BUILT USING A DISTANCE-TYPE FEATURE SPACE COMBINED WITH SMOTE RE-SAMPLING PROVIDED OPTIMAL MODEL PERFORMANCE

@@

When evaluating model performances, we found that regardless of resampling technique, random forest classifiers built using a distance-type feature space yielded consistently higher MCCs when compared to the other two predictor types (Figure 3). This indicated that models are better able to discriminate between TAD and non-TAD boundary regions when provided with spatially related associations. Delving further, we found that when combined with SMOTE re-sampling, random forest classifiers built using distance-type predictors out performed other re-sampling techniques across resolutions, suggesting that a combination of both random over- and under-sampling can improve TAD boundary prediction. Interestingly, given these parameters, we found that performances peaked at 10 kb resolution and then degraded as data resolution decreased. This reason for the improved performance between 5 kb and 10 kb data resolutions can be attributed to the added sparcity in the interaction frequency matrices at finer resolutions. This sparcity corresponds to inaccurate TAD calling which leads to worse model performance. Additionally, performances tended to be much more variable between chromosomes at lower resolutions. Moreover, this decreasing trend in performance was much more evident under the optimal combination of data level characteristics (SMOTE resampling with distance type predictors). <!--These results imply that there is an ideal resolution in TAD boundary prediction, and that that an optimally performing cell line specific random forest classifier, one that is able to best discriminate between genomic bins that are near versus far away from TAD boundaries, is one that is built on 10 kb resolution data, with distance-type predictors, and incorporates both random under- and over-sampling.--> Models implemented on the K562 cell line were found to yield the same optimal data level characteristics (Supplementary Figure 6).

![**Models built using distance-type functional genomic annotations combined with SMOTE re-sampling yielded optimal MCCs across resolution.** Barplots comparing the performances of random forest classifiers built using different combinations of data level characteristics. Within each re-sampling technique (A-D), performances are compared for each type of feature space: OC (red), OP (green), and Distance (blue), across resolutions. Performances were evaluated using MCCs and aggregated by averaging across each chromosome, with error bars representing 1 standard deviation away from the mean.](Figures/model_performances_v3_bars.png)
